Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-06-25 paper/paper.tex
revised section 8

%probably Store Widening \& Abstract Garbage Collection
\section{To Direct-Style} \label{directstyle}

@@

\subsection{Back to Direct-Style with Delimited Control} \label{uncps}

@@

\textit{Abstract Interpretation and Control-Flow Analysis.}
\citeauthor{cousot1977abstract} invented abstract interpretation as a sound approach
to approximate a program's runtime behavior\cite{cousot1977abstract}.
Control-flow analysis is one instance of abstract interpretation on functional programs
that can be traced to Jones \cite{jones1981flow}. \citeauthor{Shivers:1988:CFA:53990.54007} 
introduced $k$-CFA which uses $k$ recent calling contexts as program contours that 
differentiate values from different contexts \cite{Shivers:1988:CFA:53990.54007, 
Shivers:1991:SSC:115865.115884}.
The recent formulation of control-flow analysis is the abstracting abstract machines
methodology \cite{van2012systematic, van2010abstracting} which forms the starting point
of this paper.
Modern programming languages such as Java\cite{might2010resolving} and Racket 
\cite{Tobin-Hochstadt:2012:HSE:2384616.2384655}
can also be modeled by the abstracting abstract machine approach. \\

@@

Over the years, Danvy and his collaborators applied refunctionalization and defunctionalization 
to many different abstract machines and semantics, including the CEK machines, the CLS machines,
the SECD machines, etc.
\cite{Ager:2003:FCE:888251.888254, Danvy:2001:DW:773184.773202, danvy2004refocusing, 
Danvy:2008:DIP:1411204.1411206, AGER2004223, ager2005functional, Danvy:2006:RW:2171265.2171268,
 danvy2009towards, biernacka2009towards}
Other applications are also found to be applicable, such as Dyck word recognizer, and 
Dijkstra's shunting-yard algorithm \cite{Danvy:2006:RW:2171265.2171268}.

For deterministic languages, \citeauthor{Danvy:2008:DIP:1411204.1411206} shows
a reduction semantics is a structural operational semantics in continuation-passing style,
where the reduction context is a defunctionalized continuation \cite{Danvy:2008:DIP:1411204.1411206}.
% Danvy also shows in nondeterministic languages ...
Our work further shows that for nondeterministic languages, there is another layer of continuations
that controls the nondeterministic choices. After identifying and explicitly exposing 
(Section~\ref{linear}) that layer of nondeterministic continuations, the whole program can 
then be refunctionalized to extended continuation-passing style.

@@

CFA2 is the first solution that solves the return-flows problem~\cite{vardoulakis2010cfa2},
but CFA2 has several drawbacks: it works only on continuation-passing style programs,
and does not supports polyvariant analysis, in addition to having an exponential time
complexity.
Pushdown control-flow analysis (PDCFA) is a mechanism which maintains this precision
through the use of a Dyke state graph representing all possible stacks contained
within the unbounded-stack machine\cite{earl2012introspective, earl2010pushdown}.
Similar to PDCFA, abstracting abstract control (AAC) is another strategy for
maintaining stack precision \cite{johnson2015abstracting}.
AAC functions by utilizing continuations which are specific to both
the source and target states of a call-site transition,
which guarantees that no spurious merging will occur during returns.

@@

Our work establish the pushdown control-flow analysis through refunctionalization and a
proper caching algorithm. The call/return is naturally matched by the stack of 
the meta language. \\

\textit{Abstracting Definitional Interpreters.}
Reynold's seminal paper \emph{Definitional interpreters for higher-order programming languages} 
\cite{Reynolds:72, Reynolds:HOSC98-revisited} shows that in
definitional interpreters, the defined language can inherit properties from the defining language.
With this insight, \citeauthor{darais2017abstracting} construct abstracting
definitional interpreters which automatically inherit the pushdown control-flow
property from its defining language because the defined language simply uses the stack of the
meta-language~\cite{darais2017abstracting}. \citeauthor{darais2017abstracting}'s abstracting
definitional interpreters work on direct-style programs, and is written in monadic style. One
of the advantages of monadic abstract interpreters is modular and composable, therefore
deploying different sensitivities or features is just applying a different monad.
Prior to that, \citeauthor{Sergey:2013:MAI:2491956.2491979} show a monadic
abstract interpreter for small-step semantics \cite{Sergey:2013:MAI:2491956.2491979}.

This paper is greatly inspired by~\citeauthor{darais2017abstracting}'s work.
The difference on the surface is that ours works for A-Normal Form $\lambda$-calculus,
instead of plain $\lambda$-calculus, although our work can also be easily adapted to handle a plain
$\lambda$-calculus.
But instead of directly showing the final form of abstract definitional interpreters,
we reveal that refunctionalization plays an important role for inheriting the
stack from the metalanguage.
The target of the series of transformations in this paper is abstract definitional
interpreters, and we additionally demonstrate abstract interpreters written direct-style with
delimited control operators. 
We use the delimited control operators @shift@ and @reset@ that are implemented in Scala 
\cite{rompf2009implementing}. The correspondence of delimited control (as well as continuations) 
and monads are well-known \cite{Danvy:1990:AC:91556.91622, wadler1992essence, danvy1992representing, moggi1991notions}.

>>>>

%probably Store Widening \& Abstract Garbage Collection
\section{Back to Direct-Style} \label{directstyle}

@@

\subsection{Back to Direct-Style with Control Operators} \label{uncps}

@@

\textit{Abstract Interpretation and Control-Flow Analysis.}
\citeauthor{cousot1977abstract} invented abstract interpretation as a sound approach to
approximate a program's runtime behavior \cite{cousot1977abstract}. Control-flow analysis
is one instance of abstract interpretation on functional programs that can be traced to
Jones \cite{jones1981flow}. \citeauthor{Shivers:1988:CFA:53990.54007} introduced $k$-CFA
which uses $k$ recent calling contexts as program contours that differentiate values from
different contexts \cite{Shivers:1988:CFA:53990.54007, Shivers:1991:SSC:115865.115884}.
The recent formulation of control-flow analysis is the abstracting abstract machines
methodology \cite{van2012systematic, van2010abstracting} which forms the starting point
of this paper. Modern programming languages such as Java \cite{might2010resolving} and
Racket \cite{Tobin-Hochstadt:2012:HSE:2384616.2384655} can also be modeled by the abstracting
abstract machine approach. \\

@@

Over the years, Danvy and his collaborators applied refunctionalization and defunctionalization 
to many different abstract machines and interpreters and build functional correspondences among
them \cite{Ager:2003:FCE:888251.888254, Danvy:2001:DW:773184.773202, danvy2004refocusing,
Danvy:2008:DIP:1411204.1411206, AGER2004223, ager2005functional, Danvy:2006:RW:2171265.2171268,
danvy2009towards, biernacka2009towards}. Such semantic artifacts include the CEK machines,
the CLS machines, the SECD machines, call-by-need evaluators, monadic evaluators, etc.
Other applications are also found to be applicable, such as Dyck word recognizer, and 
Dijkstra's shunting-yard algorithm and even Quicksort algorithm \cite{Danvy:2006:RW:2171265.2171268}.

For deterministic languages, \citeauthor{Danvy:2008:DIP:1411204.1411206} shows a reduction semantics
is a structural operational semantics in continuation-passing style, where the reduction contexts
are defunctionalized (data type representing) continuations \cite{Danvy:2008:DIP:1411204.1411206}.
Not unexpectedly, the correspondence also exists for nondeterministic languages. For example,
\citeauthor{Danvy:2001:DW:773184.773202} show that refunctionalizing a regualr expressions matcher yields
its counterpart with two layer of continuations \cite{Danvy:2001:DW:773184.773202}. And the multi-layer
continuation-passing style is the underlying semantics of control operatos like @reset@ and @shift@
\cite{Danvy:1990:AC:91556.91622}.

Our work applies these ideas to the nondeterministic AAM. The additional layer of continuation controls
the nondeterministic choice over closures. After identifying and explicitly exposing (Section~\ref{linear})
that layer of nondeterministic continuations, the whole program can be refunctionalized to extended
continuation-passing style, and then back to direct-style by the use of delimited control operators.

@@

CFA2 is the first solution that solves the return-flows problem~\cite{vardoulakis2010cfa2},
but CFA2 has several drawbacks: it works only on continuation-passing style programs, and
does not support polyvariant analysis, in addition to having an exponential time complexity.
Pushdown control-flow analysis (PDCFA) is a mechanism which maintains this precision through
the use of a Dyke state graph representing all possible stacks contained within the unbounded-stack
machine \cite{earl2012introspective, earl2010pushdown}. Similar to PDCFA, abstracting abstract
control (AAC) is another strategy for maintaining stack precision \cite{johnson2015abstracting}.
AAC functions by utilizing continuations which are specific to both the source and target states
of a call-site transition, which guarantees that no spurious merging will occur during returns.

@@

Our work establish the pushdown control-flow analysis through refunctionalization and a
proper caching algorithm. The call/return is naturally matched by the call stack of the
meta-language. \\

\textit{Abstracting Definitional Interpreters.}
Reynold's seminal paper \emph{Definitional interpreters for higher-order programming languages} 
\cite{Reynolds:72, Reynolds:HOSC98-revisited} shows that in some definitional interpreters, the
defined language can inherit properties from the defining language. With this insight,
\citeauthor{darais2017abstracting} construct abstracting definitional interpreters which
automatically inherit the pushdown control-flow property from its defining language because
the defined language simply uses the call stack of the meta-language~\cite{darais2017abstracting}.
\citeauthor{darais2017abstracting}'s abstracting definitional interpreters work on direct-style
$\lambda$-calculus, and itself is written in monadic style. One of the advantages of monadic
abstract interpreters is modular and composable, therefore deploying different sensitivities
or features is just applying a different monad. Prior to that, \citeauthor{Sergey:2013:MAI:2491956.2491979}
show a monadic abstract interpreter for small-step semantics \cite{Sergey:2013:MAI:2491956.2491979}.

This paper is greatly inspired by~\citeauthor{darais2017abstracting}'s work. The difference on the
surface is that ours works for A-Normal Form $\lambda$-calculus, instead of plain $\lambda$-calculus,
although our work can also be easily adapted to handle a plain $\lambda$-calculus. But besides directly
presenting the final form of abstract definitional interpreters, we reveal that refunctionalization
plays an important role for inheriting the call stack from the meta-language. The target of the series
of transformations in this paper is abstract definitional interpreters, and we additionally demonstrate
abstract interpreters written direct-style with delimited control operators. We use the delimited control
operators @shift@ and @reset@ that are implemented in Scala \cite{rompf2009implementing}. The
correspondence of delimited control (as well as continuations) and monads are well-known
\cite{Danvy:1990:AC:91556.91622, wadler1992essence, danvy1992representing, moggi1991notions}.
